---
title: "Homework 1 – One Sample Inference"
output:
  html_document:
    code_folding: hide
---


Human toddlers, starting at around 18 months old, can learn novel object labels passively by overhearing third-party interactions without being directly addressed. For example, if one adult says “hand me an apple” and another does so, the child might learn to identify an apple and associate it with the word “apple”.


This process is driven by developing sociocognitive capacities such as gaze following, conversation monitoring, intention understanding, and perspective-taking. Of course, this process is more efficient when adults use simplified sentences, emphasis, enthusiasm, and other non-verbal emphasis.


Dogs share some human-like social and cognitive abilities they learn being immersed in our households. For example, many family dogs learn actions like “sit.” Dror et al. (2026) explored whether dogs could be capable of passive label learning, similar to toddlers. Specifically, they were interested in Gifted Word Learner (GWL) dogs that can learn the names of hundreds of objects and perform at a word-learning level comparable to 18 months olds.


To learn whether GWL dogs learn object labels by passively overhearing interactions between humans the researchers performed an experiment where 𝑛 = 10 dogs were exposed to a condition where two family members sitting across from each other passing the toy back and forth. The caretaker addressed the second person using simple sentence frames and non-verbal cues, neither directly to the dog.


This exposure occurred daily for eleven days with two new toys. 
    - Toy A: Days 1-3; 9
    - Toy B: Days 5-7; 10
    - Rest: Day 4, 8, 11

On the twelfth day, out of the caretaker’s sight, the two new toys were placed alongside nine familiar toys. The caretaker requested the toys verbally twenty-four times – twelve times for the two new toys, and twelve times for the nine familiar toys.

# Question 1

## Part a
In the paper, Dror et al. (2026) performed a Wilcoxon signed-rank test to assess whether the dogs were better than chance, which they took as 50%, for returning the new objects when requested. The 𝑛 = 10 dogs had the following results:


{1.0000, 1.0000, 0.7500, 1.0000, 0.4167, 0.8333, 0.7500, 0.8333, 0.8333, 1.0000}.


They also considered the paired difference (direct − overhearing):


{−0.0833, 0.0000, −0.0833, −0.0833, 0.5833, 0.0833, −0.2500, 0.1667, 0.0000, −0.4167}.

AI provided the following “insight” for analyzing these data: 

Report the One-Sample t-Test if your data meets basic normality assumptions, or the Wilcoxon Signed-Rank Test if you prefer not to assume normality.


Evaluate the author’s choice to use the Wilcoxon signed-rank test (not specifying exact or approximate). Also consider a one-sample 𝑡-test and a one-sample 𝑧-test (or exact binomial test). Which choices are reasonable given the data, and which choice would you have made?

**Solution:** AI recommended that if the data met normality assumptions, to then report a one-sample t-Test, however, because there are only n=10 dogs, we cannot enact the central limit theorem (CLT) to make the data behave as normal. So, in a case where the data is bound between 0 and 1 with a limited sample size, the t-Test might not be the most accurate. Moving on with the z-Test, it seems the natural choice because it deals with proportions and relies on bernoulli data. Again, while we could coerce the trails into success(1) and failure(0) for each trial a dog participates in, we have the same subject gathering data for 12 trials. This could skew the data as if one dog were inherently very good, it would be difficult to compare to another who wasn't as they are not independent trials. That leaves only the Wilcoxon signed-rank test. But the authors' failure to specify exact vs. approximate matters more than it looks. The exact version of the test requires no ties among the ranked values, and here 7 of the 10 values are tied (three dogs at 1.0000, two at 0.8333, two at 0.7500). However, I still agree with the author's use of the Wilcoxon signed-rank test as the best option.

## Part b
Perform the analyses you decided on above. Provide a full report as if you were submitting your homework to the researchers themselves. This part should be a stand alone analysis that can be read and understood independently.

Note: If you didn’t select a test in the previous part, you can use a randomization test here.


```{r}
library(ggplot2)

new.toy.prop <- c(1.0000, 1.0000, 0.7500, 1.0000, 0.4167,
                   0.8333, 0.7500, 0.8333, 0.8333, 1.0000)

dog.dat <- data.frame(dog = factor(1:10), prop.correct = new.toy.prop)

ggplot(dog.dat, aes(x = "", y = prop.correct)) +
  geom_boxplot(width = 0.3) +
  geom_jitter(width = 0.05, height = 0) +
  geom_hline(yintercept = 0.5, linetype = "dotted") +
  coord_flip() +
  labs(x = NULL, y = "Proportion of new toys retrieved") +
  theme_bw()
```
```{r}
new.toy.prop <- c(1.0000, 1.0000, 0.7500, 1.0000, 0.4167,
                   0.8333, 0.7500, 0.8333, 0.8333, 1.0000)
test.result <- wilcox.test(x = new.toy.prop, # data
mu = 0.5, # M0 = .5
alternative = "two.sided", # Ha: M != M0
conf.level = 0.95) # alpha = 0.05
test.result
```

**Question:** Did the dogs retrieve newly-labeled toys at a rate better than chance (50%) after only overhearing them named, across n = 10 dogs?
```{r}
library(dplyr)
library(e1071)
library(knitr)

new.toy.prop <- c(1.0000, 1.0000, 0.7500, 1.0000, 0.4167,
                   0.8333, 0.7500, 0.8333, 0.8333, 1.0000)

summary.tab <- tibble(prop = new.toy.prop) |>
  summarise(
    n                = n(),
    Mean             = mean(prop),
    Median           = median(prop),
    SD               = sd(prop),
    IQR              = IQR(prop),
    Skewness         = skewness(prop),
    `Excess Kurtosis` = kurtosis(prop)
  )

kable(summary.tab, digits = 2, caption = "Proportion of New Toys Retrieved (n = 10 dogs)")
```

We used a Wilcoxon signed-rank test (H₀: median proportion = 0.5) as the primary test. Because 7 of the 10 values are tied (three dogs at 1.000, two at 0.833, two at 0.750), an exact p-value isn't computable.

**Conclusion:** The dogs retrieved newly-labeled toys significantly more often than the 50% chance level after passively overhearing them named (Wilcoxon V = 54, p = .0074; sign test p = .0215, 9/10 dogs above chance), consistent with the researchers' claim of passive label learning in this sample. Given n = 10, we'd caution against over-interpreting the exact p-values — the ceiling effect (4/10 dogs at a perfect score) suggests the task may not have had enough range to fully capture variation in performance, and a slightly harder retrieval task in a larger sample would strengthen the estimate.